---
title: "Tree-based differential testing using inferential uncertainty for RNA-Seq"
author: "Noor Pratap Singh"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
output:
  rmarkdown::html_document:
    highlight: tango
    toc: true
    toc_float: true
vignette: |
  %\VignetteIndexEntry{1. Tree-based differential testing using inferential uncertainty for RNA-Seq}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(tidy=FALSE, cache=FALSE, dev="png",
                      message=FALSE, error=FALSE, warning=FALSE)
```
# Installation
```{r}
devtools::install_github("NPSDC/trenDi", "dev")
```

# trenDi
The *trenDi* method performs differential testing use the transcripts-tree obtained from *TreeTerminus*.
The transcripts are arranged in a way such that the uncertainty decreases on ascending the tree, with the
inner nodes comprising transcripts groups. *trenDi* aims to leverage the tree structure to find differential
signal in the tree which might get lost due to uncertainty. The candidate nodes can consist of both transcripts
and inner nodes, with inner nodes representing a set of transcripts that exhibit high uncertainty.

# Quick start

We demonstrate *trenDi* on a small simulated RNA-Seq dataset in the [brainSimSmall](https://github.com/NPSDC/brainSimSmall) package.
We have *Salmon* quantified counts for the samples, with 100 inferential replicates generated using Gibbs Sampling. *Note* Gibbs Sampling
can be replaced by bootstrap sampling. The dataset consists of 12 samples belonging to two-groups, with 6 samples in each.

## Loading data 
We first load the data using [beaveR](https://github.com/NPSDC/beaveR) to create *TreeSummarizedExperiment* object. 
The leaves consist of the transcripts.
```{r eval=FALSE, message=FALSE, warning=FALSE}
dir <- system.file("extdata", package="brainSimSmall")
clustFile <- file.path(dir, "cluster_nwk.txt")
quantDir <- file.path(dir, "salmon_quants")
samples <- as.vector(outer(c(1:6), c(1,2), function(x,y) paste(x,y,sep='_')))
quantFiles <- file.path(quantDir, samples, 'quant.sf')
coldata <- data.frame(files=quantFiles, names=samples, condition=factor(rep(1:2, each=6)))
tse <- beaveR::buildTSE(treeTermFile = clustFile, coldata = coldata)
print(tse)
tree <- TreeSummarizedExperiment::rowTree(tse)
print(tree)
l <- length(tree$tip)
```

## Differential transcript expression
We first perform differential testing on the transcripts using Swish. We can see the transcripts
that are called significant at the 0.01 FDR threshold.
```{r eval=FALSE, message=FALSE, warning=FALSE}
set.seed(1)
yTxps <- tse[1:l,]
yTxps <- fishpond::scaleInfReps(yTxps)
yTxps <- fishpond::labelKeep(yTxps)
yTxps <- fishpond::swish(yTxps, x="condition")
yTxps <- yTxps[match(rownames(yTxps), tree$tip.label),]
print(table(mcols(yTxps)[["qvalue"]] < 0.01))
```

## Running trenDi
We next run *trenDi* on the object *tse* created above. The pipeline is similar to Swish, where first
the data is normalized (we used median-ratio scaling) in this example. It has been broken down
into two steps, in the first we compute the size factors and then use the computed size factors for scaling the
inferential replicates. We have used *Swish* for computing the pvalues, however we compute them separately for
the leaves and the inner nodes. This is because due to lower uncertainty, the width of the distribution of the 
test-statistic of inner nodes is wider and thus have smaller p-values compared to the inner nodes.

*trenDi* outputs a set of candidate nodes based on a filtering criteria. Once a node
that has a p-value less than a certain threshold (governed by \strong(alpha)) is found, we check for
different criteria. First, nodes should consist of atleast one child node that has a p-value larger than
a threshold, otherwise we are overaggregating. We then check if all the underlying descendant nodes that have a 
confident direction of sign-change have the same direction change between conditions. The confidence is determined
by the \strong(minP) parameters and looks at the inferential replicates. Finally, we also want that the children nodes
of a candidate, tp have uncertain abundance estimates (governed by \strong(meanInfRV)).

```{r eval=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages(library(trenDi))
set.seed(1)
yAll <- beaveR::computeSizeFactors(tse)
yAll <- beaveR::scInfReps(yAll)
yAll <- fishpond::labelKeep(yAll)
metadata(yAll)$infRepsScaled <- TRUE
yInn <- yAll[(l+1):nrow(yAll),]
yInn <- fishpond::swish(yInn, x="condition")

pvals <- c(mcols(yTxps)[["pvalue"]], mcols(yInn)[["pvalue"]])
tD <- trenDi::trenDi(yAll, x="condition", pvalues = pvals,
                    minP=0.7, mIRVThresh=0.4, alpha=0.01)
print(length(tD$candNodes)) ## Total number of candidate nodes
print(sum(tD$candNodes > l)) ## Number of inner nodes that are candidates
```

## Differential gene expression
```{r eval=FALSE, message=FALSE, warning=FALSE}
gy <- tximeta::summarizeToGene(tse[1:l,])
gy <- fishpond::scaleInfReps(gy)
gy <- fishpond::labelKeep(gy)
gy <- fishpond::swish(gy, x="condition")
print(dim(gy))
print(table(mcols(gy)[["qvalue"]] < 0.01))
```